---
output:
  html_document: default
---
# Exploring the NBA versus the WNBA
As lifelong basketball fans, we all found ourselves pondering - keeping BLM in mind following the events of last year - if the NBA, a longstanding American institution, is truly free of racism - at least in the paycheck department. While exploring effects of race within the NBA, we found ourselves wondering about more information we could deduce about basketball as a profession for other _minorities_ or _underpriviledged_ sections of the society. We made our way to the WNBA (Women's NBA). We gathered data from NBA and WNBA players and compared significant information including wages. Comparing salaries across genders in the two organizations - NBA and WNBA - we found that the difference in the salaries for both genders is drastic. Women in basketball get paid less than men in millions of dollars. Through this exploratory analysis, we want to understand the reasons for the difference in salaries. Is it purely sexism? or other factors such as average player performance, time commitment required, popularity or any other factors could completely or partially explain the difference in pay.



# Summary Statistics Table
```{r include=FALSE}
library(readxl)
library(dplyr)
library(tidyverse)
library(knitr)
```

```{r, echo=FALSE}
nba_basketball_reference <- read_excel("data/nba_basketball-reference.xlsx")
wnba_basketball_reference <- read_excel("data/wnba_basketball-reference.xlsx")

basketball_info <- list()
basketball_info$num_players_nba <- nrow(nba_basketball_reference)
basketball_info$highest_average_nba_ppg <- nba_basketball_reference %>%
  filter(PTS == max(PTS, na.rm = T)) %>%
  select(PTS)
basketball_info$highest_average_nba_rebounds <- nba_basketball_reference %>%
  filter(TRB == max(TRB, na.rm = T)) %>%
  select(TRB)
basketball_info$highest_average_nba_mins <- nba_basketball_reference %>%
  filter(MP == max(MP, na.rm = T)) %>%
  select(MP)
basketball_info$highest_average_nba_freethrow_attempts <- nba_basketball_reference %>%
  filter(FTA == max(FTA, na.rm = T)) %>%
  select(FTA)
basketball_info$num_players_wnba <- nrow(wnba_basketball_reference)
basketball_info$highest_average_wnba_ppg <- wnba_basketball_reference %>%
  filter(PTS == max(PTS, na.rm = T)) %>%
  select(PTS)
basketball_info$highest_average_wnba_rebounds <- wnba_basketball_reference %>%
  filter(TRB == max(TRB, na.rm = T)) %>%
  select(TRB)
basketball_info$highest_average_wnba_mins <- wnba_basketball_reference %>%
  filter(MP == max(MP, na.rm = T)) %>%
  select(MP)
basketball_info$highest_average_wnba_freethrow_attempts <- wnba_basketball_reference %>%
  filter(FTA == max(FTA, na.rm = T)) %>%
  select(FTA)

bb_info_table <- table(basketball_info)
bb_info_table <- data.frame(bb_info_table)
renamed_table <- bb_info_table %>%
  rename(
    "Number.of.NBA.Players" = "num_players_nba",
    "Highest.Average.NBA.PPG" = "highest_average_nba_ppg",
    "Highest.Average.NBA.Rebounds" = "highest_average_nba_rebounds",
    "Highest.Average.NBA.Minutes" = "highest_average_nba_mins",
    "Highest.Average.NBA.FTAs" = "highest_average_nba_freethrow_attempts",
    "Number.of.WNBA.Players" = "num_players_wnba",
    "Highest.Average.WNBA.PPG" = "highest_average_wnba_ppg",
    "Highest.Average.WNBA.Rebounds" = "highest_average_wnba_rebounds",
    "Highest.Average.WNBA.Minutes" = "highest_average_wnba_mins",
    "Highest.Average.WNBA.FTAs" = "highest_average_wnba_freethrow_attempts",
  )
grouped_bb_table <- group_by(renamed_table, Number.of.NBA.Players, Number.of.WNBA.Players)
grouped_bb_table <- subset(grouped_bb_table, select = -c(Freq))
kable(grouped_bb_table)
```
For this table, what we wanted to was highlight was the similarity between athletes in the NBA and the WNBA. For years, it has been assumed that women are just automatically worse athletes than their male counterparts - not even coming close in terms of athletic performance. But, as we can see in the above table, this is not true. Although the highest averages of players in the NBA are higher than those in the WNBA, professional basketball is professional basketball - and all things considered, these states are fairly comparable. Of course, this then begs the question of **why** NBA players tend to make millions more than players in the WNBA - and this is a question we look forward to exploring in the final product.
