---
output:
  html_document: default
---
# Exploring the NBA versus the WNBA
Hey Shruti, for this opening I would just talk about the NBA and the WNBA and how player salaries differ so drastically between leagues. Bring up how we compiled player statistics for both leagues and how for this exploratory analysis we are using them to try and find any obvious differences between the leagues that would warrant such a massive difference in salary. 


# Summary Statistics Table
```{r include=FALSE}
library(readxl)
library(dplyr)
library(tidyverse)
library(knitr)
```

```{r, echo=FALSE}
nba_basketball_reference <- read_excel("data/nba_basketball-reference.xlsx")
wnba_basketball_reference <- read_excel("data/wnba_basketball-reference.xlsx")

basketball_info <- list()
basketball_info$num_players_nba <- nrow(nba_basketball_reference)
basketball_info$highest_average_nba_ppg <- nba_basketball_reference %>%
  filter(PTS == max(PTS, na.rm = T)) %>%
  select(PTS)
basketball_info$highest_average_nba_rebounds <- nba_basketball_reference %>%
  filter(TRB == max(TRB, na.rm = T)) %>%
  select(TRB)
basketball_info$highest_average_nba_mins <- nba_basketball_reference %>%
  filter(MP == max(MP, na.rm = T)) %>%
  select(MP)
basketball_info$highest_average_nba_freethrow_attempts <- nba_basketball_reference %>%
  filter(FTA == max(FTA, na.rm = T)) %>%
  select(FTA)
basketball_info$num_players_wnba <- nrow(wnba_basketball_reference)
basketball_info$highest_average_wnba_ppg <- wnba_basketball_reference %>%
  filter(PTS == max(PTS, na.rm = T)) %>%
  select(PTS)
basketball_info$highest_average_wnba_rebounds <- wnba_basketball_reference %>%
  filter(TRB == max(TRB, na.rm = T)) %>%
  select(TRB)
basketball_info$highest_average_wnba_mins <- wnba_basketball_reference %>%
  filter(MP == max(MP, na.rm = T)) %>%
  select(MP)
basketball_info$highest_average_wnba_freethrow_attempts <- wnba_basketball_reference %>%
  filter(FTA == max(FTA, na.rm = T)) %>%
  select(FTA)

bb_info_table <- table(basketball_info)
bb_info_table <- data.frame(bb_info_table)
renamed_table <- bb_info_table %>%
  rename(
    "Number.of.NBA.Players" = "num_players_nba",
    "Highest.Average.NBA.PPG" = "highest_average_nba_ppg",
    "Highest.Average.NBA.Rebounds" = "highest_average_nba_rebounds",
    "Highest.Average.NBA.Minutes" = "highest_average_nba_mins",
    "Highest.Average.NBA.FTAs" = "highest_average_nba_freethrow_attempts",
    "Number.of.WNBA.Players" = "num_players_wnba",
    "Highest.Average.WNBA.PPG" = "highest_average_wnba_ppg",
    "Highest.Average.WNBA.Rebounds" = "highest_average_wnba_rebounds",
    "Highest.Average.WNBA.Minutes" = "highest_average_wnba_mins",
    "Highest.Average.WNBA.FTAs" = "highest_average_wnba_freethrow_attempts",
  )
grouped_bb_table <- group_by(renamed_table, Number.of.NBA.Players, Number.of.WNBA.Players)
grouped_bb_table <- subset(grouped_bb_table, select = -c(Freq))
kable(grouped_bb_table)
```
For this table, what we wanted to was highlight was the similarity between athletes in the NBA and the WNBA. For years, it has been assumed that women are just automatically worse athletes than their male counterparts - not even coming close in terms of athletic performance. But, as we can see in the above table, this is not true. Although the highest averages of players in the NBA are higher than those in the WNBA, professional basketball is professional basketball - and all things considered, these states are fairly comparable. Of course, this then begs the question of **why** NBA players tend to make millions more than players in the WNBA - and this is a question we look forward to exploring in the final product. 
